**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**

### 2015-10-01 Schloss_Cluster_PeerJ_2015.Rmd
Editing

@@

Third, the authors attempted to describe the effects of the clustering instability on comparisons of communities. They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset. Inspection of these results indicates that although they were able to detect significant p-values, they were were of marginal biological significance. Adonis R statistics close to zero indicate the community structures from the full and subsetted datasets overlapped while values of one indicate the communities are completely different. The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL). Regardless of the statistical or biological significance of these results, the analysis does not make sense since the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering algorithms are run multiple times with different random number seeds. To overcome this, the authors selected representative sequences from OTUs and merged OTUs across different sized libraries to link OTUs. The justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable. It is not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels. Because this analysis was so poorly designed, we did not seek to reproduce it.

@@

```{r rand_ref_analysis}
rand_ref_u <- read.table(file="data/rand_ref/hits.uclosed.summary", row.names=1)
rand_ref_v <- read.table(file="data/rand_ref/hits.vclosed.summary", row.names=1)

```

***Additional problems with reference-based clustering.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs. This was unsurprising since, by definition, the algorithm represents a one-to-one mapping of reads to a reference and treats the input sequences independently. An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database. We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference algorithm with the full MiSeq dataset. Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations. On average there were `r round(rand_ref_u["mean", 1], 2)` OTUs per randomization (range from `r round(rand_ref["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`). The unrandomized reference resulted in `r rand_ref_u["original", 1]` OTUs, below the minimum observed number of OTUs when the references were randomized. This surprising result is likely because a heuristic is employed by USEARCH to determine when to stop searching the database such that the best possible match is not necessarily returned. If the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points. To test this, we substituted VSEARCH for USEARCH in the closed-reference algorithm because VSEARCH does not utilize the heuristic but is otherwise modeled after USEARCH. In fact, all randomizations of the reference sequences and the original ordering of the reference sequences generated `r rand_ref_v["original", 1]` OTUs.

@@

***Roesch dataset processed according to He study***
After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length. Although they reported observing a total of 50,542 sequences that were represented by 13,293 unique sequences, we obtained a total of `r length(scan("data/he/canada_soil.good.unique.pick.redundant.fasta", what="", sep="\n", quiet=T))/2` sequences that were represented by `r length(scan("data/he/canada_soil.good.unique.pick.fasta", what="", sep="\n", quiet=T))/2` unique sequences. Similar to the He study, we randomly sampled, without replacement, 20, 40, 60, and 80% of the sequences from the full data set. The random sampling was repeated 30 times. The order of the sequences in the full dataset was randomly permuted without replacement to generate an additional 30 datasets. For the hierarchical clustering algorithms and to generate a distance matrix, the pairwise distances between sequences were calculated in mothur using the pairwise.seqs command with the default Needleman-Wunsch alignment algorithm and parameters. Execution of the clustering algorithms was performed as described in the original He study using mothur and QIIME as appropriate.

>>>>

```{r knitr.settings, eval=TRUE, echo=TRUE, cache=TRUE}
opts_chunk$set("fig.path"="figure/")
opts_chunk$set("fig.align"="center")
opts_chunk$set("dev" = c("png", "pdf"))

opts_chunk$set("tidy" = TRUE)

opts_chunk$set("echo" = TRUE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)
```

@@

Third, the authors attempted to describe the effects of the clustering instability on comparisons of communities. They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset. Inspection of these results indicates that although they were able to detect significant p-values, they were were of marginal biological significance. Adonis R statistics close to zero indicate the community structures from the full and subsetted datasets overlapped while values of one indicate the communities are completely different. The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL). Regardless of the statistical or biological significance of these results, the analysis does not make sense since the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering algorithms are run multiple times with different random number seeds. To overcome this, the authors selected representative sequences from each OTU and used those representative sequences to link OTU assignments between the different sized sequence sets. The justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable. It is not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels. Because this analysis was so poorly designed, we did not seek to reproduce it.

@@

```{r rand_ref_analysis}
rand_ref_u <- read.table(file="data/rand_ref/hits.uclosed.summary", row.names=1)
rand_ref_v <- read.table(file="data/rand_ref/hits.vclosed.summary", row.names=1)
```

***; reference-based clustering.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs. This was unsurprising since, by definition, the algorithm represents a one-to-one mapping of reads to a reference and treats the input sequences independently. An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database. We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference algorithm with the full MiSeq dataset. Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations. On average there were `r round(rand_ref_u["mean", 1], 2)` OTUs per randomization (range from `r round(rand_ref["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`). The unrandomized reference resulted in `r rand_ref_u["original", 1]` OTUs, below the minimum observed number of OTUs when the references were randomized. This surprising result is likely because a heuristic is employed by USEARCH to determine when to stop searching the database such that the best possible match is not necessarily returned. If the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points. To test this, we substituted VSEARCH for USEARCH in the closed-reference algorithm because VSEARCH does not utilize the heuristic but is otherwise modeled after USEARCH. In fact, all randomizations of the reference sequences and the original ordering of the reference sequences generated `r rand_ref_v["original", 1]` OTUs.

@@

***Roesch Canadian soil dataset***
After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length. Although they reported observing a total of 50,542 sequences that were represented by 13,293 unique sequences, we obtained a total of `r length(scan("data/he/canada_soil.good.unique.pick.redundant.fasta", what="", sep="\n", quiet=T))/2` sequences that were represented by `r length(scan("data/he/canada_soil.good.unique.pick.fasta", what="", sep="\n", quiet=T))/2` unique sequences. Similar to the He study, we randomly sampled, without replacement, 20, 40, 60, and 80% of the sequences from the full data set. The random sampling was repeated 30 times. The order of the sequences in the full dataset was randomly permuted without replacement to generate an additional 30 datasets. For the hierarchical clustering algorithms and to generate a distance matrix, the pairwise distances between sequences were calculated in mothur using the pairwise.seqs command with the default Needleman-Wunsch alignment algorithm and parameters. Execution of the clustering algorithms was performed as described in the original He study using mothur (v.1.37) and QIIME (v.1.XX) as appropriate.
